A Transdiciplinary approach to developing a spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities in and around the coast of South Africa: Combining ecology 

### 2019-08-21 literature_review.tex
sections under waves edited and added to. simple word output kept but template added for later use. pdf output failed. Will attempt to fix.

---
title: "A Transdiciplinary approach to developing a spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities in and around the coast of South Africa: Combining ecology and 3D hyperspectral hydrodynamic modelling."
author: "Ross Coppin"
output: 
  word_document: 
    fig_caption: yes
    toc: yes
---

```{r include=FALSE}
library(png)
library(grid)
```

# Summary

Marine ecosystems are maintained by a variety of complex interactions between abiotic and biotic variables such as temperature, wave exposure, pH, competition, and processes such as top-down and bottom-up control, predator-prey relationships and phenology (Doney et al. 2012; Harley et al. 2006; Burrows et al. 2011; McGowan et al. 1998). These abiotic and biotic variables, the interactions between them, and the various ecological processes, ultimately determine the community composition and ecological functioning of all ecosystems (Guimaraes & Coutinho 1996; Harley et al. 2006; Poloczanska et al. 2013; Jennings & Brander 2010; Polovina 2005; Johnson et al. 2011; Wernberg et al. 2016; Dayton et al. 1999). Climate directly and indirectly affects the way in which abiotic and biotic variables interact, but is often compounded by other impacts such as habitat destruction, pollution, and overfishing (Blamey et al. 2012; Blamey et al. 2015). Temperature and wave exposure have been recognised as important variables with regards to climate-driven changes within the ocean (Guimaraes & Coutinho 1996; McGowan et al. 1998; McQuaid & Branch 1984; Laufkötter et al. 2015; Belkin 2009; Filbee-Dexter et al. 2018; Miller et al. 2011; Smale et al. 2011; Smale & Moore 2017). In order to persist and survive within variable and changing environments, organisms must either migrate, adapt, or die.

Seaweeds are sessile organisms which are unable to migrate to new areas when local environmental conditions become unsuitable and therefore are forced to adapt to new conditions in order to avoid expiration (Dayton 1985; Tegner et al. 1997). The main form of mortality for seaweeds is through mechanical dislodgement by wave action (Schiel et al. 2006; Seymour et al. 1989; Thomsen et al. 2004; Cavanaugh et al. 2011; Graham et al. 1997; Demes et al. 2013; Edwards & Estes 2006). Seaweeds, particularly brown seaweeds, are able to undergo rapid morphological adaptation to the hydrodynamic environment (Miller et al. 2011; Friedland & Denny 1995; Wing et al. 2007; Wernberg & Thomsen 2005; Fowler-Walker et al. 2006; Dudgeon & Johnson 1992; Blanchette et al. 2002). This allows seaweeds to reduce mortality through mechanical dislodgement by inducing morphology which reduces overall drag. Seaweeds that are unable to avoid mechanical dislodgement either raft out to sea or wash up onto beaches. Not all beach-cast kelp may have originated from a nearby kelp population and may have originated from other sites or regions of the coast through rafting and ocean currents (Malm et al. 2004; Emery & Tschudy 1941; Filbee-Dexter & Scheibling 2012; Hobday 2000). Therefore, because kelp morphology is specific to its local environment the morphological features may be able to indicate, within a certain amount of probability, what site or region it most likely originated from. In other words, beach-cast kelp may be used as proxy for investigating the flow of coastal currents. Using kelp as a proxy for determining its original location will be calibrated by means of a hydrodynamic model which will be designed from already existing SWAN and Delft3D models. This combined approach will allow investigation into flow regimes around the west and south-west coasts of South Africa and the role they play in subtidal ecology. Furthermore, the model could be used to determine transport of plastic pollution around the coast. For instance, micro-plastics are recognised as a threat to marine life (Seltenrich 2015; Azzarello & Van Vleet 1987; Wright et al. 2013; Derraik 2002; Andrady 2011; Fendall & Sewell 2009; Ivar Do Sul & Costa 2014; O ’donoghue & Marshall n.d.; Ballent et al. 2012), however very little is known about how microplastics may be transported along the coast.

The increased use of plastic in society over the past half century has resulted in large amounts of plastic litter in both the marine and terrestrial environment (Andrady 2011; Wright et al. 2013; Cole et al. 2011). The problems associated with large plastic debris have received attention for many decades, whereas those connected to marine micro-plastics comparatively received very little attention. However, today it has become a prioritized area among political organizations, agencies and NGOs around the world. The micro-plastic debris present in the ocean are derived from marine and terrestrial sources (Derraik 2002; Wright et al. 2013; Cole et al. 2011; Seltenrich 2015), however there is little understanding of how microplastics may be distributed with ocean currents. Therefore, the coupling the adaptability of kelp morphology and the ability to simulate hydrodynamic processes can greatly improve our understanding of transport pathways and likely locations of accumulation. This in turn may inform management decisions with regards to eliminating and managing marine pollution in South Africa.

# Background

@@

Temperature and wave exposure have been shown to be important drivers of seaweed distribution, physiological functioning, ability to recover, population dynamics and morphology. Mechanical forces generated by the hydrodynamic environment, in the form of sudden strong ocean currents or storms, between 10- 20 m s-1 with accelerations of 400 m s-2 (Friedland and Denny 1995) are the biggest threat to kelp survival. Kelps are able to rapidly adapt their morphological characteristics to reduce drag and avoid dislodgement (Blanchette 1997). For example a study by Koehl et al. (2008) showed that transplanted Nereocystis luetkeana plants from a wave sheltered site to a wave exposed site changed their morphology to flat blades and narrow laterals that are less prone to drag forces in 4-5 days. Another study by Fowler-Walker, Wernberg, and Connell (2006) tested for differences in morphology of Ecklonia radiata between wave-sheltered and wave-exposed sites and through a combination of in situ sampling and transplantation of juvenile plants. The results showed that morphology differed between wave-sheltered and wave-exposed sites (thin thallus at sheltered sites and a narrow, thick thallus with a thick stipe at exposed sites), and was consistent with previous studies. Juveniles transplanted into wave exposed sites under went rapid morphological adaptation, whilst the opposite was true for wave-sheltered sites which showed slower morphological adaptation.

@@

Advances in numerical modelling has gained much traction in recent years and has been applied in a variety of ways with regards to ecological studies. For example, a study by Wang and Xia (2009) used the Delft3D-Flow model to assess the hydraulic suitability of a stream as a spawning ground for the Chinese Sturgeon (Acipenser sinensis) in the Yangtze River. The authors calculated the horizontal mean vorticity which was used to assess the hydraulic environment of spawning ground. The flow field state was determined through model simulation and field-measured data used to validate the model. The results added to existing scientific database for spawning ground hydraulic environmental protection. Different numerical models can often be integrated to model across ecosystem levels. For example a study by Leon et al. (2003) used integrated physical (Delft3D hydrological model) and bio-chemical (Agricultural Non-point Source model) processes models to investigate the possible impact on the Lake Malawi water quality due to management actions performed at the watershed level.

Since wave energy is an important driver in marine ecosystems, particularly kelp, the advances in hydrodynamic modelling offer a new opportunity for multifactorial and quantitative approach to research in marine ecosystems. The Delft3D and SWAN models have been used successfully in previous studies regarding brine plume discharge, impacts of storms, effects of climate change on the hydrological environment etc. The models have not been designed for shallow environments (<6m) and therefore may not be suitable to model coastal hydrological environments. However these models may be adjusted to suit coastal waters if they are combined with a new numerical model which can be calibrated to suit these needs.

In recent years there has been growing attention on plastic pollution, particularly in the ocean. Plastic pollution can be in the form of macro- and microplastics. Microplastics are tiny plastic granules used as scrubbers in cosmetics and air-blasting, and small plastic fragments that originate from larger pieces of plastic known as macroplastics, while macroplastics…insert definition here… The potential harms of of plastic pollution in the marine environment was highlighted in the 1970’s and renewed interest has lead to research showing that plastic pollution in the ocean are widespread. Plastics may become bio-available to biota in the food-web which may cause problems with an organism's physiological functioning. Furthermore, the relatively large surface area and composition of microplastics provides an environment that is able of adhering to organic pollutants. In other words microplastics also act as a vector for transport and assimilation of organic pollutants.

@@

# Kelp environmental drivers

Important environmental drivers of kelp individuals and communities include light, substrata, salinity, sedimentation, nutrients, temperature and wave exposure. Although studies have investigated the effects of these important environmental drivers, the roles these factors play is often difficult to evaluate as such factors may never be fully independent of each other, i.e. environmental factors are to some extent are dependent on one another. Multifactorial studies have attempted to explain combined effects, however these studies are often limited to investigating combination of two or three environmental drivers as inclusion of too many factors can lead to results that are difficult to interpret. Environmental factors are highly variable on temporal and spatial scales and their effects may also be dependent on the life-stage of the organism, adding a further layer of complexity to investigations.

## Light

@@

## Substrata

## Salinity

## Depth

Depth does not affect kelp ecosystems directly, however a change in depth causes fluctuations or changes in other environmental variables such as water motion, light and temperature. Water motion also decreases with depth, and some kelps better suited to deeper environments (*L. pallida*) replace those in the shallows (*E. maxima*) (Dayton 1985; Gerard 1982). The increase in depth can lead to a decrease in sunlight penetration, with some species better adapted for low-light conditions than others, such as (*L. pallida*). Temperature may also change along a depth gradient due to a reduction in sunlight penetration (Dayton 1985; Gerard 1982). Therefore depth does not directly play a role in kelp functioning but may alter more influential factors such as light and water motion.

## Sedimentation

## Nutrients

@@

# Ocean and coastal waves

## Introduction

Regardless of the location around the world, waves are a feature of any coastline and the marine environment, and are important manifestations of energy in the ocean. Waves are not the movement of water particles but rather the movement and propagation of energy through the ocean. The source of this energy can be formed locally, such as wind-driven waves, and or from distant locations in the ocean such as storms, know as *swell*. This energy is transferred from deeper water into shallower water where it plays a role in driving complex marine ecosystems as well as shaping the environments that they live in. Energy that is left-over after these processes is transferred into heat energy, and heats up the sand and rocks on which each wave hits. This energy can also be harnessed in simple ways such as a surfer catching a wave, or more complex ways such as capturing energy from the ocean environment for electricity production. However, waves are stochastic in nature and therefore the energy that propagates through the ocean is not always consistent. Therefore, it is often difficult to quantify and predict wave energy in the marine environment. Despite this fact, waves are gaining more recognition for the role they play in shaping coastlines, beaches and hence the communities and organisms that depend on these systems.

Waves manifest themselves in different ways, which is also dependent on the energy creating force, and can be classified into different categories. For instance, "chop" are produced by local winds, while “tsunamis”" are rare waves that are formed during a earthquakes or landslides and can be produced from thousands of kilometers away. Waves in the ocean are known as “swell” and are produced from far distant storms, while the most consistent form of waves that interacts with the coastlines, usually twice a day, are called “tides”. Tides are produced from differences in gravitational forces between the ocean, earth’s crust, sun and the moon. These classifications can be broken down further and will be investigated later in this chapter. The focus of this chapter will be understanding wave theory and how waves form and propagate into shallow-coastal waters. This will be essential to later chapters that aim to model wave energy and calibrate the produced model with biotic variables.

## Generating and restoring forces

Waves are formed due to the constant interaction between two forces, which are known as the "generating force" and the “restoration force”. The generating force is the force which pushes water from one layer up into the other. Layers in the ocean are formed through differences in temperature (thermocline) and salinity (pycnocline) which creates a boundry for wave energy to move along. In other words, waves occur along the boundaries formed in the ocean by various abiotic processes, and therefore density boundaries are essential to the propagation of energy through the ocean.

The generating force and restoring force occur along these boundaries, the generating force pushes water up across ocean boundaries while the restoring force pulls it back to where the boundary was originally, trying to restore the balance in energy. This "tug of war" between the two forces creates an oscillating motion between the boundary layers which acts as a point of disturbance, sending out energy in all directions. The disturbance energy will continue to manifest itself provided the tug and pull between the two forces is occuring. Once the generating force stops, and ultimately the energy from the point of disturbance, the waves dissipate and water restores to its original state. A simple example would be blowing on water in a cup. The blowing of air onto the surface of the water in the cup creates a point of disturbance that pushes the air boundary into the water boundary. The restoring force is the surface tension of the water that is maintained through hydrogen bonds between water molecules. Once blowing on the water in the cup has stopped, the ripples or “waves” in the cup begin to diminish in size until the surface tension returns to its original state. The hydrogen bonds between the water molecules are stronger than the force of gravity, and therefore the force of the surface tension returns the water to its original state. These waves are known as “capillary waves” and is essentially residual energy after the generating source has stopped. The restoring force can also be in the form of gravity and are known as “gravity waves”. Using the same example, if one blows too hard, the water boundary is pushed up into the air boundary, breaking the hydrogen bonds between water molecules which allows gravity to return the water to its original state. All waves in the ocean are either capillary waves or gravity waves, and their classification will be dependent on the restoring force involved.

@@

## Wave physics and scales

@@

## Types of waves

There are a variety of waves that form in the ocean and all differ in terms of period or wavelength (??image??). The longest wave that can form in the ocean are known as trans-tidal waves, and are generated by fluctuations in magnetism between the Earth’s crust and atmosphere. The magnetic push and pull between the Moon and the Sun creates waves with a slightly shorter wavelength, known as tides. Their period and wavelength can also range from a few hours to more than a day, and from a few hundred to a thousand kilometers respectively. Storm surge tend to have a slightly shorter wavelength and period compared to tides. When low atmospheric pressure systems and high wind speeds in a storm it elevates the ocean surface, generating storm surge, which may cause flooding in coastal areas as it approaches the coastline. Tsunami’s are on the lower end of the scale, and are generated by earthquakes or submarine ‘landslides. Their random nature makes them difficult to predict and increase amplitude as they approach the coast. This can make them considerably large along coastal areas and often cause immense damage and loss of life.

## Measuring waves

Waves are often thought of as an elevation of the sea surface from a specific point over a period of time but this is obvisiously not he case. This is known as *surface elevation* and is the instaneous elevation of the sea surface abouve a specific point in a time record (see figure ???). Although surface elevation does not represent a wave, it can be used to create a wave profile. This is achieved by profiling the the surface elevation between two successive *downward zero-crossings* or *upwardward zero-crossings* (see figure ??). Both zero-crossings are symmetrical and essentially the same statistically. However, in practice the downward zero-crossings are preferred as it takes steepness of a wave into account (the front, see figure ??) which is relevant to characterising breaking waves. It should be noted that surface elevation can be negative while a wave profile cannot.

```{r surface elevation figures, echo=FALSE, warning=FALSE, fig.cap="The definition of a wave in a time record of the surface elevation with downward zero-crossings (upper panel) and upward zero-crossings (lower panel)."}

surface_elevation <- readPNG("figures/surface_elevation_waves.png")
grid.raster(surface_elevation)
```

@@

```{r Hs_Tp_surface_elevation figure, echo=FALSE, warning=FALSE, fig.cap="The definition of wave height and wave period in a time record of the surface elevation (the wave is defined with downwrd zero-crossings."}

Hs_Tp_elevation <- readPNG("figures/Hs_Tp_surface_elevation.png")
grid.raster(Hs_Tp_elevation)
```

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={A Transdiciplinary approach to developing a spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities in and around the coast of South Africa: Combining ecology and 3D hyperspectral hydrodynamic modelling.},
            pdfauthor={Ross Coppin},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{A Transdiciplinary approach to developing a spectral numerical
coastal-flow model based on passive macroalgal rafting phenenomea to
shed light on ambient and stochastic dispersal and connectivity of
nearshore and offshore marine communities in and around the coast of
South Africa: Combining ecology and 3D hyperspectral hydrodynamic
modelling.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Ross Coppin}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

\begin{document}
\maketitle

\hypertarget{summary}{%
\section{Summary}\label{summary}}

Marine ecosystems are maintained by a variety of complex interactions
between abiotic and biotic variables such as temperature, wave exposure,
pH, competition, and processes such as top-down and bottom-up control,
predator-prey relationships and phenology (Doney et al.~2012; Harley et
al.~2006; Burrows et al.~2011; McGowan et al.~1998). These abiotic and
biotic variables, the interactions between them, and the various
ecological processes, ultimately determine the community composition and
ecological functioning of all ecosystems (Guimaraes \& Coutinho 1996;
Harley et al.~2006; Poloczanska et al.~2013; Jennings \& Brander 2010;
Polovina 2005; Johnson et al.~2011; Wernberg et al.~2016; Dayton et
al.~1999). Climate directly and indirectly affects the way in which
abiotic and biotic variables interact, but is often compounded by other
impacts such as habitat destruction, pollution, and overfishing (Blamey
et al.~2012; Blamey et al.~2015). Temperature and wave exposure have
been recognised as important variables with regards to climate-driven
changes within the ocean (Guimaraes \& Coutinho 1996; McGowan et
al.~1998; McQuaid \& Branch 1984; Laufkötter et al.~2015; Belkin 2009;
Filbee-Dexter et al.~2018; Miller et al.~2011; Smale et al.~2011; Smale
\& Moore 2017). In order to persist and survive within variable and
changing environments, organisms must either migrate, adapt, or die.

Seaweeds are sessile organisms which are unable to migrate to new areas
when local environmental conditions become unsuitable and therefore are
forced to adapt to new conditions in order to avoid expiration (Dayton
1985; Tegner et al.~1997). The main form of mortality for seaweeds is
through mechanical dislodgement by wave action (Schiel et al.~2006;
Seymour et al.~1989; Thomsen et al.~2004; Cavanaugh et al.~2011; Graham
et al.~1997; Demes et al.~2013; Edwards \& Estes 2006). Seaweeds,
particularly brown seaweeds, are able to undergo rapid morphological
adaptation to the hydrodynamic environment (Miller et al.~2011;
Friedland \& Denny 1995; Wing et al.~2007; Wernberg \& Thomsen 2005;
Fowler-Walker et al.~2006; Dudgeon \& Johnson 1992; Blanchette et
al.~2002). This allows seaweeds to reduce mortality through mechanical
dislodgement by inducing morphology which reduces overall drag. Seaweeds
that are unable to avoid mechanical dislodgement either raft out to sea
or wash up onto beaches. Not all beach-cast kelp may have originated
from a nearby kelp population and may have originated from other sites
or regions of the coast through rafting and ocean currents (Malm et
al.~2004; Emery \& Tschudy 1941; Filbee-Dexter \& Scheibling 2012;
Hobday 2000). Therefore, because kelp morphology is specific to its
local environment the morphological features may be able to indicate,
within a certain amount of probability, what site or region it most
likely originated from. In other words, beach-cast kelp may be used as
proxy for investigating the flow of coastal currents. Using kelp as a
proxy for determining its original location will be calibrated by means
of a hydrodynamic model which will be designed from already existing
SWAN and Delft3D models. This combined approach will allow investigation
into flow regimes around the west and south-west coasts of South Africa
and the role they play in subtidal ecology. Furthermore, the model could
be used to determine transport of plastic pollution around the coast.
For instance, micro-plastics are recognised as a threat to marine life
(Seltenrich 2015; Azzarello \& Van Vleet 1987; Wright et al.~2013;
Derraik 2002; Andrady 2011; Fendall \& Sewell 2009; Ivar Do Sul \& Costa
2014; O 'donoghue \& Marshall n.d.; Ballent et al.~2012), however very
little is known about how microplastics may be transported along the
coast.

The increased use of plastic in society over the past half century has
resulted in large amounts of plastic litter in both the marine and
terrestrial environment (Andrady 2011; Wright et al.~2013; Cole et
al.~2011). The problems associated with large plastic debris have
received attention for many decades, whereas those connected to marine
micro-plastics comparatively received very little attention. However,
today it has become a prioritized area among political organizations,
agencies and NGOs around the world. The micro-plastic debris present in
the ocean are derived from marine and terrestrial sources (Derraik 2002;
Wright et al.~2013; Cole et al.~2011; Seltenrich 2015), however there is
little understanding of how microplastics may be distributed with ocean
currents. Therefore, the coupling the adaptability of kelp morphology
and the ability to simulate hydrodynamic processes can greatly improve
our understanding of transport pathways and likely locations of
accumulation. This in turn may inform management decisions with regards
to eliminating and managing marine pollution in South Africa.

\hypertarget{background}{%
\section{Background}\label{background}}

@@

Temperature and wave exposure have been shown to be important drivers of
seaweed distribution, physiological functioning, ability to recover,
population dynamics and morphology. Mechanical forces generated by the
hydrodynamic environment, in the form of sudden strong ocean currents or
storms, between 10- 20 m s-1 with accelerations of 400 m s-2 (Friedland
and Denny 1995) are the biggest threat to kelp survival. Kelps are able
to rapidly adapt their morphological characteristics to reduce drag and
avoid dislodgement (Blanchette 1997). For example a study by Koehl et
al.~(2008) showed that transplanted Nereocystis luetkeana plants from a
wave sheltered site to a wave exposed site changed their morphology to
flat blades and narrow laterals that are less prone to drag forces in
4-5 days. Another study by Fowler-Walker, Wernberg, and Connell (2006)
tested for differences in morphology of Ecklonia radiata between
wave-sheltered and wave-exposed sites and through a combination of in
situ sampling and transplantation of juvenile plants. The results showed
that morphology differed between wave-sheltered and wave-exposed sites
(thin thallus at sheltered sites and a narrow, thick thallus with a
thick stipe at exposed sites), and was consistent with previous studies.
Juveniles transplanted into wave exposed sites under went rapid
morphological adaptation, whilst the opposite was true for
wave-sheltered sites which showed slower morphological adaptation.

@@

Advances in numerical modelling has gained much traction in recent years
and has been applied in a variety of ways with regards to ecological
studies. For example, a study by Wang and Xia (2009) used the
Delft3D-Flow model to assess the hydraulic suitability of a stream as a
spawning ground for the Chinese Sturgeon (Acipenser sinensis) in the
Yangtze River. The authors calculated the horizontal mean vorticity
which was used to assess the hydraulic environment of spawning ground.
The flow field state was determined through model simulation and
field-measured data used to validate the model. The results added to
existing scientific database for spawning ground hydraulic environmental
protection. Different numerical models can often be integrated to model
across ecosystem levels. For example a study by Leon et al.~(2003) used
integrated physical (Delft3D hydrological model) and bio-chemical
(Agricultural Non-point Source model) processes models to investigate
the possible impact on the Lake Malawi water quality due to management
actions performed at the watershed level.

Since wave energy is an important driver in marine ecosystems,
particularly kelp, the advances in hydrodynamic modelling offer a new
opportunity for multifactorial and quantitative approach to research in
marine ecosystems. The Delft3D and SWAN models have been used
successfully in previous studies regarding brine plume discharge,
impacts of storms, effects of climate change on the hydrological
environment etc. The models have not been designed for shallow
environments (\textless{}6m) and therefore may not be suitable to model
coastal hydrological environments. However these models may be adjusted
to suit coastal waters if they are combined with a new numerical model
which can be calibrated to suit these needs.

In recent years there has been growing attention on plastic pollution,
particularly in the ocean. Plastic pollution can be in the form of
macro- and microplastics. Microplastics are tiny plastic granules used
as scrubbers in cosmetics and air-blasting, and small plastic fragments
that originate from larger pieces of plastic known as macroplastics,
while macroplastics\ldots{}insert definition here\ldots{} The potential
harms of of plastic pollution in the marine environment was highlighted
in the 1970's and renewed interest has lead to research showing that
plastic pollution in the ocean are widespread. Plastics may become
bio-available to biota in the food-web which may cause problems with an
organism's physiological functioning. Furthermore, the relatively large
surface area and composition of microplastics provides an environment
that is able of adhering to organic pollutants. In other words
microplastics also act as a vector for transport and assimilation of
organic pollutants.

@@

\hypertarget{kelp-environmental-drivers}{%
\section{Kelp environmental drivers}\label{kelp-environmental-drivers}}

Important environmental drivers of kelp individuals and communities
include light, substrata, salinity, sedimentation, nutrients,
temperature and wave exposure. Although studies have investigated the
effects of these important environmental drivers, the roles these
factors play is often difficult to evaluate as such factors may never be
fully independent of each other, i.e.~environmental factors are to some
extent are dependent on one another. Multifactorial studies have
attempted to explain combined effects, however these studies are often
limited to investigating combination of two or three environmental
drivers as inclusion of too many factors can lead to results that are
difficult to interpret. Environmental factors are highly variable on
temporal and spatial scales and their effects may also be dependent on
the life-stage of the organism, adding a further layer of complexity to
investigations.

\hypertarget{light}{%
\subsection{Light}\label{light}}

@@

\hypertarget{substrata}{%
\subsection{Substrata}\label{substrata}}

\hypertarget{salinity}{%
\subsection{Salinity}\label{salinity}}

\hypertarget{depth}{%
\subsection{Depth}\label{depth}}

Depth does not affect kelp ecosystems directly, however a change in
depth causes fluctuations or changes in other environmental variables
such as water motion, light and temperature. Water motion also decreases
with depth, and some kelps better suited to deeper environments
(\emph{L. pallida}) replace those in the shallows (\emph{E. maxima})
(Dayton 1985; Gerard 1982). The increase in depth can lead to a decrease
in sunlight penetration, with some species better adapted for low-light
conditions than others, such as (\emph{L. pallida}). Temperature may
also change along a depth gradient due to a reduction in sunlight
penetration (Dayton 1985; Gerard 1982). Therefore depth does not
directly play a role in kelp functioning but may alter more influential
factors such as light and water motion.

\hypertarget{sedimentation}{%
\subsection{Sedimentation}\label{sedimentation}}

\hypertarget{nutrients}{%
\subsection{Nutrients}\label{nutrients}}

@@

\hypertarget{ocean-and-coastal-waves}{%
\section{Ocean and coastal waves}\label{ocean-and-coastal-waves}}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Regardless of the location around the world, waves are a feature of any
coastline and the marine environment and are important manifestations of
energy in the ocean. Waves are not the movement of water particles but
is rather the movement and propagation of energy through the ocean. The
source of this energy can be formed locally, such as wind-driven waves,
and or from distant locations in the ocean, such as storms. This energy
is then transferred from deeper water into shallower water where it
plays a role in driving beach morphology. Sand and rock are eroded with
each passing wave, thereby playing a role in shaping coastlines. Energy
that is left-over after these processes is transferred into heat energy,
and heats up the sand and rocks on which each wave hits. This energy can
also be harnessed in simple ways such as a surfer catching a wave, or
more complex ways such as capturing energy from the ocean environment
for electricity production. However, waves are stochastic in nature and
therefore the energy that propagates through the ocean is not always
consistent. Therefore, it is often difficult to quantify and predict
wave energy in the marine environment. Despite this fact, waves are
gaining more recognition for the role they play in shaping coastlines,
beaches and hence the communities and organisms that depend on these
systems.

Waves manifest themselves in different ways, which is also dependent on
the energy creating force, and can be classified into different
categories. For instance, ``chop'' are produced by local winds, while
``tsunamis''" are rare waves that are formed during a earthquakes or
landslides and can be produced from thousands of kilometers away. Waves
in the ocean are known as ``swell'' and are produced from far distant
storms, while the most consistent form of waves that interacts with the
coastlines, usually twice a day, are called ``tides''. Tides are
produced from differences in gravitational forces between the ocean,
earth's crust, sun and the moon. These classifications can be broken
down further and will be investigated later in this chapter. The focus
of this chapter will be understanding wave theory and how waves form and
propagate into shallow-coastal waters. This will be essential to later
chapters that aim to model wave energy and calibrate the produced model
with biotic variables.

\hypertarget{generating-and-restoring-forces}{%
\subsection{Generating and restoring
forces}\label{generating-and-restoring-forces}}

Waves are formed due to the constant interaction between two forces,
which are known as the ``generating source'' and the ``restoration
force''. The generating force is the force which pushes water from one
layer up into the other, for example pushing water up across the
boundary of air. This can also occur in the ocean layers. The ocean
layers through differences in density, which in turn is partly driven by
abiotic parameters such as temperature and salinity, which creates a
boundary for wave energy to move along. Examples of such boundaries are
along the ocean surface, pycnoclines, and the sea-floor. optional
addition of pycnocline details. In other words, waves occur along the
boundaries formed in the ocean by various abiotic processes, and
therefore density boundaries are essential to the propagation of energy
through the ocean.

The generating force and restoring force occur along these boundaries,
the generating force pushes water up across ocean boundaries while the
restoring force pulls it back to where the boundary was originally,
trying to restore the balance in energy. This ``tug of war'' between the
two forces creates an oscillating motion between the boundary layers
which acts as a point of disturbance, sending out energy in all
directions. The disturbance energy will continue to manifest itself
provided the tug and pull between the two forces is occuring. Once the
generating force stops, and ultimately the energy from the point of
disturbance, the waves dissipate and water restores to its original
state. A simple example would be blowing on water in a cup. The blowing
of air onto the surface of the water in the cup creates a point of
disturbance that pushes the air boundary into the water boundary. The
restoring force is the surface tension of the water that is maintained
through hydrogen bonds between water molecules. Once blowing on the
water in the cup has stopped, the ripples or ``waves'' in the cup begin
to diminish in size until the surface tension returns to its original
state. The hydrogen bonds between the water molecules are stronger than
the force of gravity, and therefore the force of the surface tension
returns the water to its original state. These waves are known as
``capillary waves'' and is essentially residual energy after the
generating source has stopped. The restoring force can also be in the
form of gravity and are known as ``gravity waves''. Using the same
example, if one blows too hard, the water boundary is pushed up into the
air boundary, breaking the hydrogen bonds between water molecules which
allows gravity to return the water to its original state. All waves in
the ocean are either capillary waves or gravity waves, and their
classification will be dependent on the restoring force involved.

@@

\hypertarget{wave-physics-and-scales}{%
\subsection{Wave physics and scales}\label{wave-physics-and-scales}}

@@

\hypertarget{types-of-waves}{%
\subsection{Types of waves}\label{types-of-waves}}

There are a variety of waves that form in the ocean and all differ in
terms of period or wavelength (??image??). The longest wave that can
form in the ocean are known as trans-tidal waves, and are generated by
fluctuations in magnetism between the Earth's crust and atmosphere. The
magnetic push and pull between the Moon and the Sun creates waves with a
slightly shorter wavelength, known as tides. Their period and wavelength
can also range from a few hours to more than a day, and from a few
hundred to a thousand kilometers respectively. Storm surge tend to have
a slightly shorter wavelength and period compared to tides. When low
atmospheric pressure systems and high wind speeds in a storm it elevates
the ocean surface, generating storm surge, which may cause flooding in
coastal areas as it approaches the coastline. Tsunami's are on the lower
end of the scale, and are generated by earthquakes or submarine
`landslides. Their random nature makes them difficult to predict and
increase amplitude as they approach the coast. This can make them
considerably large along coastal areas and often cause immense damage
and loss of life.

\hypertarget{measuring-waves}{%
\subsection{Measuring waves}\label{measuring-waves}}

Waves are often thought of as an elevation of the sea surface from a
specific point over a period of time but this is obvisiously not he
case. This is known as \emph{surface elevation} and is the instaneous
elevation of the sea surface abouve a specific point in a time record
(see figure ???). Although surface elevation does not represent a wave,
it can be used to create a wave profile. This is achieved by profiling
the the surface elevation between two successive \emph{downward
zero-crossings} or \emph{upwardward zero-crossings} (see figure ??).
Both zero-crossings are symmetrical and essentially the same
statistically. However, in practice the downward zero-crossings are
preferred as it takes steepness of a wave into account (the front, see
figure ??) which is relevant to characterising breaking waves. It should
be noted that surface elevation can be negative while a wave profile
cannot.

\begin{figure}
\centering
\includegraphics{literature_review_files/figure-latex/surface elevation figures-1.pdf}
\caption{The definition of a wave in a time record of the surface
elevation with downward zero-crossings (upper panel) and upward
zero-crossings (lower panel).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{literature_review_files/figure-latex/Hs_Tp_surface_elevation figure-1.pdf}
\caption{The definition of wave height and wave period in a time record
of the surface elevation (the wave is defined with downwrd
zero-crossings.}
\end{figure}
